---
title: "Lab 5 - ANCOVA"
author: "Fred Clavel"
date: "February 12, 2019"
output:
  html_document:
      includes:
          before_body: header.html
          after_body: footer.html
---
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
div.green { background-color:#e6ffd0; border-radius: 5px; padding: 20px;}
div.red { background-color:#ffe6f0; border-radius: 5px; padding: 20px;}
div.warn { background-color:#ee1111; color:#ffffff; font-weight:bold; border-radius: 5px; padding: 20px;}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Setting up ANCOVA in R

ANCOVA can involved mixed effects or not, but they are always nested designs. We simply collapse the design into an analysis that involves the two highest order levels.

```{r}
d <- read.csv(file="Coral_interaction_experiment.csv", header=T, sep=",")
library(DT)
datatable(d)

```

<br>
<div class = "warn">

PROTIP: You should only use `aov` for balanced designs. For **unbalanced designs** you should use `lm`, `lme`,  or `nlme` procedures. As an alternative, you can also use the `anova` function and specify that you do NOT want to run a Type I (i.e., balanced design) ANOVA.

</div>
<br>

```{r}
#Specifying and using mixed effects:

#use the nlme package for an unbalanced design:
library(nlme)

#this specifies that size is the outcome, predicted by site, with a random effect of transect)
#model_EXAMPLE <- lme(fixed=size~site, random=~1|tran, data=d)

#test these data using a similar approach
model2 <- lme(fixed=growth~treatment, 
              random=~1|tank, data=d)
summary(model2)
```
<br>
We can also test the normality of groups assumption using the shapiro test. You should run this test at the **lowest level of grouping of your data.**  You can use the `tapply` function to run a test based on some grouping fairly quickly.
<br>
```{r}
#conduct a shapiro test on growth, grouped by the list of tank values.

tapply(d$growth, list(d$tank), shapiro.test)
```

