---
title: "Stats Lab 6 - ANCOVA"
author: "Fred Clavel"
date: "February 19, 2019"
output:
  html_document:
      includes:
          before_body: header.html
          after_body: footer.html

---
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
div.green { background-color:#e6ffd0; border-radius: 5px; padding: 20px;}
div.red { background-color:#ffe6f0; border-radius: 5px; padding: 20px;}
div.warn { background-color:#ee1111; color:#ffffff; font-weight:bold; border-radius: 5px; padding: 20px; font-variant:small-caps}
div.safe { background-color:#11dd11; color:#ffffff; font-weight:bold; border-radius: 5px; padding: 20px; font-variant:small-caps}
div.data {background-color:#ddddee; border-radius: 5px; padding: 20px; font-family:monospace}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F)
```

Conducting ANCOVA in R.

```{r}
#packages mounting




```

<div class="warn">
###SCENARIO: Mars Attacks! 

We need to know whether martians are hostile or not. Theory - tone of communication relates to the hostility metric, but the intelligence level is a linear covariate of the hostility of the martians.

We also want to see if this relationship holds up over 3 time points.

We need an ANCOVA to see if this theory pans out.

</div>


When running ANCOVA, try to identify each of the pieces. above we have

response variable - hostility
explanatory variable - tone of communication
covariate - intelligence

This will help you lay out the model more intuitively.

```{r}
#using fake data set that we do not have.
#mars <-read.csv("martians.csv", header=T)


#make sure data look ok
str(mars)
#tone is numeric - change to factor.
mars$tone <-as.factor(mars$tone)
head(mars) #check the first six rows
dim(mars)  #check the dimensions of the data set
```



STEPS FOR ANCOVA
```{r}


#step 1 is to test for an interaction. If we have an interaction, there is no point in testing for an ANCOVA with main effects.
#you can just run aov with the interaction only - it will still generate main effects with T1 SS, but the interaction is handled the same way.
model1 <- aov(time1~intelligence*tone, data=mars)

#step 2 explore the residuals
#try the outliers package

# turn residuals into a data frame
### d <-data.frame(res=Res1, tone=mars$tone)

# run cochran.test for outlying variance
### cochran.test(res~tone)

#if you save the results of the initial ANCOVA to an object called model1, you can plot the entire thing, and R will give you diagnostics.
### plot(model1)


#step 3 
###  run multiple anovas across the three time points, treating them as an independent study.

# IN homework 3, q3, run:
# aov(START~biomass*tissue, data=kelp)

################ NEVER DO THIS IN PRACTICE, UNLESS IT IS PURELY DIAGNOSTIC -
####### TIME CANNOT BE TREATED AS A NUISANCE VARIABLE IN A LONGITUDINAL MODEL
####### MOREOVER, THESE ARE MULTIPLE TESTS THAT ARE NOT ADJUSTED FOR TYPE 1 ERROR INFLATION

# each time, test for interaction first. If it is significant, there is no need to run model version b, because the main effects become irrelevant
model1a <-aov(Time1~tone * intelligence, data = mars)
model1b <-aov(Time1~tone + intelligence, data = mars)



model2a <-aov(Time2~tone * intelligence, data = mars)
model2b <-aov(Time2~tone + intelligence, data = mars)

model3a <-aov(Time3~tone * intelligence, data = mars)
model3b <-aov(Time3~tone + intelligence, data = mars)


```

